---
title: "Plant Height Model"
description: "Three-parameter logistic model fitted to plant height as a function of accumulated growing degree-days (GDD), with ANOVA on model parameters and all derivative-based critical points."
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  cache    = FALSE,
  comment  = "##",
  collapse = TRUE,
  warning  = FALSE,
  message  = FALSE
)
```

## Overview

The same logistic framework applied to leaf area is used here for **plant height**
(stem length, cm). Parameters β₁–β₃ are estimated independently for each
season × cultivar × block combination and subjected to two-factor ANOVA in
randomised complete blocks (FAT2DBC). Critical points derived from the first and
second derivatives are also computed at the block level and analysed by ANOVA.

$$
y = \frac{\beta_1}{1 + e^{\;\beta_3(\beta_2 - x)}}
$$

| Parameter | Interpretation |
|-----------|---------------|
| β₁ | Upper asymptote (maximum plant height, cm) |
| β₂ | Inflection point (GDD at maximum elongation rate) |
| β₃ | Steepness coefficient (elongation rate) |

## Required packages

```{r libraries}
#| cache: false
library(rio)        # Data import
library(tidyverse)  # Data wrangling and visualisation
library(metan)      # mean_by(), doo()
library(broom)      # tidy() – NLS summary as data frame
library(AgroR)      # FAT2DBC() – two-factor ANOVA in RCBD
library(patchwork)  # Combine ggplot panels
library(hydroGOF)   # gof() – goodness-of-fit metrics
library(gt)         # Professional tables
```

## Load data

```{r load-data}
# Import the pre-computed plot-mean dataset.
# Each row = one season × cultivar × block × date mean (already averaged across plants).
# This dataset is used both for model fitting and for the ANOVA on critical points.
df_model <-
  import("../df_model_cresc_medias.xlsx")
```

## Fit logistic models – plant height

```{r fit-nls}
# Response variable: 'cp' (comprimento de planta = plant height, cm)
formula_ap <- cp_mean ~ b1 / (1 + exp((b3 * (b2 - gda))))

# Starting values:
#   b1 = 100  → expected maximum height (~100 cm)
#   b2 = 400  → inflection point around 400 GDD
#   b3 = 0.01 → moderate steepness
start_ap <- list(b1 = 100, b2 = 400, b3 = 0.01)

# Fit one model per season × cultivar × block combination.
# The date "04/11" falls outside the vegetative growth window (post-flowering)
# and is excluded to preserve the sigmoid shape of the curve.
mod_ap <-
  df_model |>
  filter(data != "04/11") |>
  group_by(epoca, cultivar, bloco) |>
  doo(~nls(formula_ap, data = ., start = start_ap))
```

## Goodness of fit

```{r goodness-of-fit}
#| cache: false
# Helper function: extract R² and AIC from a single NLS model object
get_gof <- function(model) {
  aic  <- AIC(model)
  bic  <- BIC(model)
  fit  <- model$m$fitted()
  res  <- model$m$resid()
  obs  <- fit + res
  
  g <- hydroGOF::gof(sim = fit, obs = obs, digits = 4)
  
  r2   <- g["R2", ]
  rmse <- g["RMSE", ]
  mae  <- g["MAE", ]
  d    <- g["d", ]
  
  data.frame(aic = aic, bic = bic, r2 = r2, rmse = rmse, mae = mae, d = d)
}

# Apply to all models
gof_ap <-
  mod_ap |>
  mutate(map_dfr(.x = data, .f = ~get_gof(.))) |>
  select(-data) |>
  mutate(variable = "plant_height")

# Display goodness of fit table with gt
gof_ap |> 
  gt() |> 
  tab_header(
    title = "Supplementary Table S4",
    subtitle = "Goodness-of-fit metrics for Plant Height logistic models (per block)"
  ) |> 
  fmt_number(columns = where(is.numeric), decimals = 4)

saveRDS(gof_ap, "../results/gof_ap.rds")
```

## Extract estimated parameters and compute critical points

```{r extract-params}
# broom::tidy() converts each NLS object into a tidy data frame;
# pivot_wider() places b1, b2, b3 as separate columns.
parameters_ap <-
  mod_ap |>
  mutate(data = map(data, ~.x |> tidy())) |>
  unnest(data) |>
  select(epoca, cultivar, bloco, term, estimate) |>
  pivot_wider(names_from  = term,
              values_from = estimate)
```

## Derivative functions

```{r derivative-functions}
# Symbolically verify the first derivative
D(expression(b1 / (1 + exp((b3 * (b2 - x))))), "x")

# First derivative: instantaneous elongation rate at a given GDD (x)
dy <- function(x, b1, b2, b3) {
  b1 * (exp((b3 * (b2 - x))) * b3) / (1 + exp((b3 * (b2 - x))))^2
}

# Symbolically verify the second derivative
D(expression(b1 * (exp((b3 * (b2 - x))) * b3) / (1 + exp((b3 * (b2 - x))))^2), "x")

# Second derivative: rate of change of elongation rate (acceleration / deceleration)
d2y <- function(x, b1, b2, b3) {
  -(b1 * (exp((b3 * (b2 - x))) * b3 * b3) / (1 + exp((b3 * (b2 - x))))^2 -
      b1 * (exp((b3 * (b2 - x))) * b3) *
      (2 * (exp((b3 * (b2 - x))) * b3 * (1 + exp((b3 * (b2 - x)))))) /
      ((1 + exp((b3 * (b2 - x))))^2)^2)
}
```

### Critical points at block level

```{r critical-points}
# Compute ALL critical points at the individual block level.
# This preserves the replication structure required for ANOVA.
#
# Critical points derived from the logistic model:
#   xpi  = β₂                              → GDD at maximum elongation rate (inflection)
#   ypi  = dy(β₂)                          → maximum elongation rate value
#   xmap = β₂ + ln(2 + √3) / β₃           → GDD at maximum acceleration point (MAP)
#   xmdp = β₂ − ln(2 + √3) / β₃           → GDD at maximum deceleration point (MDP)
#   ymap = d²y(xmap)                        → second-derivative value at MAP
#   ymdp = d²y(xmdp)                        → second-derivative value at MDP
parameters_ap <-
  parameters_ap |>
  mutate(
    xpi  = b2,
    ypi  = dy(xpi, b1, b2, b3),
    xmap = b2 - (log(2 + sqrt(3)) / b3),
    xmdp = b2 + (log(2 + sqrt(3)) / b3),
    ymap = d2y(xmap, b1, b2, b3),
    ymdp = d2y(xmdp, b1, b2, b3)
  ) |>
  as.data.frame() |>
  mutate(variable = "plant_height")

# Block-level means for derivative plots (used in stat_function args)
parameters_ap_mean <-
  parameters_ap |>
  metan::mean_by(epoca, cultivar)

# Factor-level means (Sowing Season and Cultivar)
# Since interaction is often non-significant, we present main effects
means_epoca_ap <- 
  parameters_ap |> 
  metan::mean_by(epoca) |> 
  mutate(Factor = "Sowing Season") |> 
  rename(Level = epoca)

means_cultivar_ap <- 
  parameters_ap |> 
  metan::mean_by(cultivar) |> 
  mutate(Factor = "Cultivar") |> 
  rename(Level = cultivar)

means_summary_ap <- 
  bind_rows(means_epoca_ap, means_cultivar_ap) |> 
  relocate(Factor, Level)



# Display means summary using gt
means_summary_ap |> 
  gt(groupname_col = "Factor") |> 
  tab_header(
    title = "Supplementary Table S5", 
    subtitle = "Main Effect Means: Model parameters and critical points for Plant Height"
  ) |> 
  fmt_number(columns = where(is.numeric), decimals = 5)

saveRDS(means_summary_ap, "../results/means_summary_ap.rds")

```

## ANOVA on model parameters and critical points

```{r anova-params}
#| cache: false
# List of variables to analyze
vars_ap <- c("b1", "b2", "b3", "xpi", "ypi", "xmap", "xmdp")

# 1. Run the ANOVA for all variables and store the full objects in a list
# This allows access to means tests and other FAT2DBC outputs later.
mod_anova_ap <- 
  vars_ap |> 
  set_names() |> 
  map(~with(parameters_ap, FAT2DBC(epoca, cultivar, bloco, get(.x))))

# 2. Extract and combine tidy ANOVA tables for the summary
anova_ap <- 
  mod_anova_ap |> 
  map_dfr(~{
    .x$anova |> 
      broom::tidy()
  }, .id = "Parameter") |> 
  mutate(term = str_replace(term, "Fator1", "Season"),
         term = str_replace(term, "Fator2", "Cultivar"))

# Display the summary table with gt
anova_ap |> 
  filter(term != "Residuals", Parameter != "xpi") |> 
  gt() |> 
  tab_header(
    title = "Supplementary Table S6",
    subtitle = "ANOVA Summary: Model parameters and critical points for Plant Height"
  ) |> 
  fmt_number(columns = where(is.numeric), decimals = 5) |> 
  tab_style(
    style = cell_text(color = "red", weight = "bold"),
    locations = cells_body(
      rows = p.value < 0.05
    )
  )

saveRDS(anova_ap, "../results/anova_ap.rds")
```

## Fitted logistic curves

```{r fitted-plot, fig.cap="Logistic curves fitted to plant height (cm) as a function of accumulated GDD, faceted by sowing season. Points = treatment means. The semi-transparent point (04/11) was excluded from fitting.", fig.width=6, fig.height=10}
# Logistic formula in x/y notation (required by geom_smooth internals)
formula_xy <- y ~ b1 / (1 + exp((b3 * (b2 - x))))

ggplot(
  df_model |> filter(!data %in% c("04/11")),
  aes(gda, cp_mean, color = cultivar)
) +
  geom_smooth(
    method      = "nls",
    method.args = list(
      formula = formula_xy,
      start   = c(b1 = 60, b2 = 400, b3 = 0.01)
    ),
    se  = FALSE,
    aes(color = cultivar)
  ) +
  facet_wrap(~epoca, ncol = 1) +
  stat_summary(
    fun      = mean,
    geom     = "point",
    aes(color = cultivar),
    size     = 2.5,
    position = position_dodge(width = 0.8)
  ) +
  scale_y_continuous(breaks = seq(0, 150, by = 25)) +
  labs(
    x     = "Accumulated Degree Days",
    y     = "Plant height (cm)",
    color = "Cultivar"
  ) +
  scale_color_manual(values = c("gold", "brown")) +
  # Excluded point shown transparently for reference
  stat_summary(
    data        = df_model |> filter(data == "04/11"),
    aes(x = gda, y = cp_mean),
    fun         = mean,
    alpha       = 0.4,
    shape       = 16,
    show.legend = FALSE
  ) +
  theme_bw(base_size = 18) +
  theme(
    panel.grid.major  = element_blank(),
    legend.background = element_rect(fill = "transparent"),
    legend.position   = "bottom"
  )

ggsave("../figs/mod_logistico_ap.jpg", width = 6, height = 9)
```

## Derivative plots

```{r derivative-plots, fig.cap="First (left) and second (right) derivatives of the fitted logistic curves for plant height. Points mark the inflection (●), maximum acceleration (▲), and maximum deceleration (▽) thresholds per cultivar × season combination.", fig.width=8, fig.height=10}

# --- First derivative plot ---
plot_pi_ap <-
  ggplot() +
  stat_function(fun = dy, aes(color = "D", linetype = "E1"), n = 500, linewidth = 1,
                xlim = c(0, 1000),
                args = c(b1 = parameters_ap_mean[[1, 3]],
                         b2 = parameters_ap_mean[[1, 4]],
                         b3 = parameters_ap_mean[[1, 5]])) +
  stat_function(fun = dy, aes(color = "M", linetype = "E1"), linewidth = 1,
                xlim = c(0, 1000),
                args = c(b1 = parameters_ap_mean[[2, 3]],
                         b2 = parameters_ap_mean[[2, 4]],
                         b3 = parameters_ap_mean[[2, 5]])) +
  stat_function(fun = dy, aes(color = "D", linetype = "E2"), linewidth = 1,
                xlim = c(0, 1000),
                args = c(b1 = parameters_ap_mean[[3, 3]],
                         b2 = parameters_ap_mean[[3, 4]],
                         b3 = parameters_ap_mean[[3, 5]])) +
  stat_function(fun = dy, aes(color = "M", linetype = "E2"), linewidth = 1,
                xlim = c(0, 1000),
                args = c(b1 = parameters_ap_mean[[4, 3]],
                         b2 = parameters_ap_mean[[4, 4]],
                         b3 = parameters_ap_mean[[4, 5]])) +
  # Mark the inflection point for each treatment
  geom_point(aes(xpi, ypi, shape = epoca, color = cultivar),
             data = parameters_ap_mean, size = 3, show.legend = FALSE) +
  theme_bw(base_size = 20) +
  scale_x_continuous(breaks = seq(0, 1200, by = 200)) +
  labs(
    x        = "Accumulated Degree Days",
    y        = expression(Plant ~ height ~ (cm ~ plant^{-1} ~ degree ~ C ~ day^{-1})),
    color    = "Cultivar",
    linetype = "Season"
  ) +
  theme(legend.position = "bottom", panel.grid.minor = element_blank()) +
  scale_color_manual(values = c("gold", "brown"))

# --- Second derivative plot ---
plot_sd_ap <-
  ggplot() +
  geom_hline(yintercept = 0) +   # zero line: above = acceleration, below = deceleration
  stat_function(fun = d2y, aes(color = "D", linetype = "E1"), n = 500, linewidth = 1,
                xlim = c(0, 1000),
                args = c(b1 = parameters_ap_mean[[1, 3]],
                         b2 = parameters_ap_mean[[1, 4]],
                         b3 = parameters_ap_mean[[1, 5]])) +
  stat_function(fun = d2y, aes(color = "M", linetype = "E1"), n = 500, linewidth = 1,
                xlim = c(0, 1000),
                args = c(b1 = parameters_ap_mean[[2, 3]],
                         b2 = parameters_ap_mean[[2, 4]],
                         b3 = parameters_ap_mean[[2, 5]])) +
  stat_function(fun = d2y, aes(color = "D", linetype = "E2"), n = 500, linewidth = 1,
                xlim = c(0, 1000),
                args = c(b1 = parameters_ap_mean[[3, 3]],
                         b2 = parameters_ap_mean[[3, 4]],
                         b3 = parameters_ap_mean[[3, 5]])) +
  stat_function(fun = d2y, aes(color = "M", linetype = "E2"), n = 500, linewidth = 1,
                xlim = c(0, 1000),
                args = c(b1 = parameters_ap_mean[[4, 3]],
                         b2 = parameters_ap_mean[[4, 4]],
                         b3 = parameters_ap_mean[[4, 5]])) +
  # MAP (triangle up) and MDP (triangle down) markers
  geom_point(aes(xmap, ymap, fill = cultivar), data = parameters_ap_mean,
             size = 3, shape = 24, show.legend = FALSE) +
  geom_point(aes(xmdp, ymdp, fill = cultivar), data = parameters_ap_mean,
             size = 3, shape = 25, show.legend = FALSE) +
  theme_bw(base_size = 20) +
  labs(
    x        = "Accumulated Degree Days",
    y        = expression(Plant ~ height ~ (cm ~ plant^{-1} ~ degree ~ C ~ day^{-2})),
    color    = "Cultivar",
    linetype = "Season"
  ) +
  theme(legend.position = "bottom", panel.grid.minor = element_blank()) +
  scale_color_manual(values = c("gold", "brown")) +
  scale_fill_manual(values  = c("gold", "brown"))

plot_pi_ap / plot_sd_ap +
  plot_layout(guides = "collect") &
  theme(legend.position = "bottom")

ggsave("../figs/deriv_ap.jpg", width = 8, height = 12)
```


# Correlation between model parameters
```{r}
# Correlation between parameters and critical points

  parameters_ap |> 
  select(all_of(vars_ap)) |> 
  metan::corr_coef() |> 
  plot()
```
