---
title: "Correlation Analysis"
description: "Pearson correlation matrix among leaf area, leaf number, and plant height across all measurement dates and treatments."
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  cache    = FALSE,
  comment  = "##",
  collapse = TRUE,
  warning  = FALSE,
  message  = FALSE
)
```

## Overview

This section examines the linear associations among the three vegetative growth
traits measured at each harvest:

- **Leaf area per plant** (`af_planta`, cm² plant⁻¹)
- **Leaf number per plant** (`n`, leaves plant⁻¹)
- **Plant height** (`cp`, cm)

A Pearson correlation matrix with significance stars is produced via
`metan::corr_plot()`.

## Required packages

```{r libraries}
library(rio)    # Data import
library(metan)  # corr_plot() – correlation matrix with visual output
library(ggplot2)
```

## Correlation plot

```{r corr-plot, fig.cap="Pearson correlation matrix among leaf area, leaf number, and plant height. The lower panel shows scatter plots with a linear smoother; the upper panel shows the correlation coefficient and its significance (*** p < 0.001, ** p < 0.01, * p < 0.05).", fig.width=6, fig.height=6}
import("../df_model_cresc.xlsx") |>
  # Select the three growth variables; corr_plot() handles pairwise correlations
  corr_plot(af_planta, n, cp)

ggsave("../figs/correlacao.jpg", width = 6, height = 6)
```

::: {.callout-note}
All three traits are expected to be positively and strongly correlated given that
they are all driven by the same thermal-time accumulation process.
:::

## Session information

```{r session-info}
sessionInfo()
```
