---
title: "Supplementary Tables"
subtitle: "Complete statistical summary for the vegetative growth of flax"
format: html
---

```{r setup, include=FALSE}
library(gt)
library(dplyr)
```

This page contains all supplementary tables cited in the main manuscript, including goodness-of-fit metrics, treatment means, and Analysis of Variance (ANOVA) for all evaluated vegetative traits.

::: {.panel-tabset}

## Leaf Area

### Goodness of Fit
```{r gof-af}
readRDS("../results/gof_af.rds") |> 
  gt() |> 
  tab_header(
    title = "Supplementary Table S1",
    subtitle = "Goodness-of-fit metrics for Leaf Area logistic models (per block)"
  ) |> 
  fmt_number(columns = where(is.numeric), decimals = 4)
```

### Main Effect Means
```{r means-af}
readRDS("../results/means_summary_af.rds") |> 
  gt(groupname_col = "Factor") |> 
  tab_header(
    title = "Supplementary Table S2", 
    subtitle = "Main Effect Means: Model parameters and critical points for Leaf Area"
  ) |> 
  fmt_number(columns = where(is.numeric), decimals = 5)
```

### ANOVA Summary
```{r anova-af}
readRDS("../results/anova_af.rds") |> 
  filter(term != "Residuals", Parameter != "xpi") |> 
  gt() |> 
  tab_header(
    title = "Supplementary Table S3",
    subtitle = "ANOVA Summary: Model parameters and critical points for Leaf Area"
  ) |> 
  fmt_number(columns = where(is.numeric), decimals = 5) |> 
  tab_style(
    style = cell_text(color = "red", weight = "bold"),
    locations = cells_body(
      rows = p.value < 0.05
    )
  )
```

## Plant Height

### Goodness of Fit
```{r gof-ap}
readRDS("../results/gof_ap.rds") |> 
  gt() |> 
  tab_header(
    title = "Supplementary Table S4",
    subtitle = "Goodness-of-fit metrics for Plant Height logistic models (per block)"
  ) |> 
  fmt_number(columns = where(is.numeric), decimals = 4)
```

### Main Effect Means
```{r means-ap}
readRDS("../results/means_summary_ap.rds") |> 
  gt(groupname_col = "Factor") |> 
  tab_header(
    title = "Supplementary Table S5", 
    subtitle = "Main Effect Means: Model parameters and critical points for Plant Height"
  ) |> 
  fmt_number(columns = where(is.numeric), decimals = 5)
```

### ANOVA Summary
```{r anova-ap}
readRDS("../results/anova_ap.rds") |> 
  filter(term != "Residuals", Parameter != "xpi") |> 
  gt() |> 
  tab_header(
    title = "Supplementary Table S6",
    subtitle = "ANOVA Summary: Model parameters and critical points for Plant Height"
  ) |> 
  fmt_number(columns = where(is.numeric), decimals = 5) |> 
  tab_style(
    style = cell_text(color = "red", weight = "bold"),
    locations = cells_body(
      rows = p.value < 0.05
    )
  )
```

## Leaf Number

### Goodness of Fit
```{r gof-nf}
readRDS("../results/gof_nf.rds") |> 
  gt() |> 
  tab_header(
    title = "Supplementary Table S7",
    subtitle = "Goodness-of-fit metrics for Leaf Number logistic models (per block)"
  ) |> 
  fmt_number(columns = where(is.numeric), decimals = 4)
```

### Main Effect Means
```{r means-nf}
readRDS("../results/means_summary_nf.rds") |> 
  gt(groupname_col = "Factor") |> 
  tab_header(
    title = "Supplementary Table S8", 
    subtitle = "Main Effect Means: Model parameters and critical points for Leaf Number"
  ) |> 
  fmt_number(columns = where(is.numeric), decimals = 5)
```

### ANOVA Summary
```{r anova-nf}
readRDS("../results/anova_nf.rds") |> 
  filter(term != "Residuals", Parameter != "xpi") |> 
  gt() |> 
  tab_header(
    title = "Supplementary Table S9",
    subtitle = "ANOVA Summary: Model parameters and critical points for Leaf Number"
  ) |> 
  fmt_number(columns = where(is.numeric), decimals = 5) |> 
  tab_style(
    style = cell_text(color = "red", weight = "bold"),
    locations = cells_body(
      rows = p.value < 0.05
    )
  )

```

:::
